Under-approximation of reachability through static analysis in multivalued asynchronous networks

### 2013-04-01 parts/def-ph.tex
typo
section 2.3

\subsection{Modeling cooperation}
%\towrite{Sortes coopératives avec mise à jour prioritaire (pas de décalage temporel)}
Cooperation between processes to make another process bounce can be expressed in PH by building a \emph{cooperative sort}, as described in \cite{PMR10-TCSB}.
\pref{fig:ph-livelock} shows an example of cooperation between processes $a_1$ and $b_1$ to make $c_0$ bounce to $c_1$:
a cooperative sort $ab$ is defined with 4 processes (one for each sub-state of the presence of processes $a_1$ and $b_1$).
For the sake of clarity, the processes of $ab$ are indexed using the sub-state they represent.
Hence, $ab_{01}$ represents the sub-state $\PHstate{a_0,b_1}$, and so on.
Each process of sort $a$ and $b$ hit $ab$ to make it bounce to the process reflecting the status of the sorts $a$ and $b$
(e.g., $\PHfrappe{a_1}{ab_{00}}{ab_{10}}$ and $\PHfrappe{a_1}{ab_{01}}{ab_{11}}$).
Then, to represent the cooperation between $a_1$ and $b_1$, the process $ab_{11}$ hits $c_0$ to make it bounce to $c_1$ instead of independent hits from $b_1$ and $a_1$.

@@

\todo{Nouvelle section pour les hypothèses + conséquences ?}
\subsection{Hypothesis} \towrite{Reformuler}
\label{ssec:hypothesis}

In the scope of this paper, we focus on a specific class of Process Hitting models.
We define here the restrictions that lead to this class of models \todo{and show that they are equivalent to discrete networks}.

\begin{criterion}[2 priorities]
\label{cr:2prio}
%\paragraph{2 priorities}
  In this paper, we only consider Process Hitting models with 2 priorities.
\end{criterion}

\pref{cr:2prio} allows to models two kinds of actions: controllable actions that model the non-determinacy of biological evolutions,
and uncontrollable actions that are used to model non-biological behaviours in the model.

In \pref{def:component} we define a well-formed component as a sort that is hit only by actions of priority $2$.
\begin{definition}
\label{def:component}
  A sort $a \in \PHs$ is said to be a well-formed component if and only if:
  $$\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 2 \enspace.$$
  We call $\components$ the set of well-formed components.
\end{definition}

\begin{definition}
\label{def:chainsactions}
  The set of chains of actions $\PHh(b, a)$ between two sorts $b$ and $a$ is defined as below:
  \begin{align*}
  \PHh(b, a) = \{ (h_i)_{i \in \segm{0}{s}} \in (\PHh^{(1)})^{s+1} &\mid s \in \sN, \PHsort(\PHhitter(h_0)) = b \wedge \PHsort(\PHtarget(h_s)) = a \\
    & \wedge \forall i \in \segm{1}{s}, \PHsort(\PHtarget(h_{i-1})) = \PHsort(\PHhitter(h_{i})) \}
  \end{align*}
\end{definition}

\begin{criterion}[Cycle-freeness between actions of priority $1$]
\label{cr:cyclefreeness}
%\paragraph{Cycle-freeness between actions of priority $1$}
  There is no cycle in chains of actions:
  $$\forall a \in \PHs, \PHh(a, a) = \emptyset \enspace.$$
\end{criterion}

Given \pref{cr:cyclefreeness}, we define the neighbour \towrite{revoir terminologie : neighbors ? vicinity ?} sorts and actions (\pref{def:neighbors}).
\begin{definition}[Neighbour sorts ($\Vs$) and Neighbour actions ($\Vh$)]
\label{def:neighbors}
  The set of neighbour sorts $\Vs(a)$ of a given sort $a$ is the set of components that can interact with it through actions of priority $1$.
  The set of neighbour actions $\Vh(a)$ of $a$ is the set of actions of priority $1$ influencing $a$.
  $$\Vs(a) = \{ b \in \components \mid \PHh(b, a) \neq \emptyset \}$$
%  $$\Vh(a) = \{ h \in \PHh^{(1)} \mid \PHsort(\PHhitter(h)) \in \Vs(a) \wedge \PHsort(\PHtarget(h)) \in \Vs(a) \cup \{ a \} \}$$
  $$\Vh(a) = \{ h \in \PHh^{(1)} \mid \exists b \in \Vs(a), \exists hs \in \PHh(b,a), h \in hs \}$$
\end{definition}

We denote $\pfp_s(a)$ the set of processes of sort $a$ that can be reached from state $s$ and that are local steady-states.
%the set of possible local steady-states of a sort $a$ from a state $s$.
%\todo{À corriger : état global ?}
\begin{definition}[Local steady-state ($\pfp$)]
\label{def:pfp}
  For all $a \in \PHa$ and $s \in \PHl$,
  \begin{align*}
    \pfp_s(a) = \{ a_i \in \PHl_a &\mid \exists \delta \in \restriction{\Sce}{1}(s), \PHget{(s \PHplay \delta)}{a} = a_i \\
    & \quad \wedge \forall \PHhit{b_i}{c_j}{c_k} \in \Vh(a), \PHget{(s \PHplay \delta)}{b} \neq b_i \vee \PHget{(s \PHplay \delta)}{c} \neq c_j \}
  \end{align*}
  Of course, if $a \in \components$, then $\pfp_s(a) = \{ \PHget{s}{a} \}$.
\end{definition}

In \pref{def:cs} we define a well-formed cooperative sort as a sort that is only by actions of priority $1$ and so that
any sub-state of its neighbourhood is represented by exactly one local steady state in this sort,
and all of its processes represents at least one sub-state of its neighbourhood. \towrite{reformuler ou supprimer}
\begin{definition}
\label{def:cs}
  A sort $a \in \PHs$ is said to be a well-formed cooperative sort iff:
  \begin{enumerate}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 1$
%    \item $\forall s \in \underset{b \in \Vs(a)}{\times} \PHl_b, |\pfp_s(a)| = 1$
%    \item $\forall a_i \in \PHl_a, \exists s \in \underset{b \in \Vs(a)}{\times} \PHl_b, \pfp_s(a) = \{ a_i \}$
    \item \label{csai} $\forall s \in \PHl, |\pfp_s(a)| = 1$
    \item \label{css} $\forall a_i \in \PHl_a, \exists s \in \PHl, \pfp_s(a) = \{ a_i \}$
  \end{enumerate}
\end{definition}
We call $\cs$ the set of well-formed cooperative sorts.
%We note that: $\forall a \in \components, \Vs(a) = \emptyset \wedge \Vh(a) = \emptyset$
Because of (\ref{csai}), we denote in the following: $\pfp_s(a) = a_i$.
Furthermore, because of (\ref{css}), we denote $\csState(a_i)$ the set of sub-states represented by the process $a_i$ of the cooperative sort $a$ (\pref{def:csState}).
\begin{definition}[Represented sub-states ($\csState$)]
\label{def:csState}
  $$\csState(a_i) = \{ ps \in \underset{b \in \Vs(a)}{\times} \PHl_b \mid \forall s \in \PHl, \pfp_{s \Cap ps}(a) = a_i \}$$
\end{definition}

\begin{criterion}[Components \& cooperative sorts]
\label{cr:compcs}
%\paragraph{Components \& cooperative sorts}
  The studied PH is a PH with 2 priorities, and it is only composed of well-formed components and cooperative sorts:
  $$\components \cup \cs = \PHs \wedge \components \cap \cs = \emptyset \enspace.$$
\end{criterion}

\begin{criterion}[Bounded termination]
\label{cr:bounded}
%\paragraph{Bounded termination}
  The dynamics of model $\restriction{\PH}{1}$ contains no cycles:
  %Cycles are still allowed in $\PH$ however.
  %Formally :
  $$\exists N \in \sN, \forall \delta \in \restriction{\Sce}{1}, |\indexes{\delta}| \leq N \enspace.$$
\end{criterion}

>>>>

\subsection{Modelling cooperation}
\label{ssec:cooperation}
%\towrite{Sortes coopératives avec mise à jour prioritaire (pas de décalage temporel)}
Cooperation between processes to make another process bounce can be expressed in PH by building a \emph{cooperative sort}, as described in \cite{PMR10-TCSB}.
\pref{fig:ph-livelock} shows an example of cooperation between processes $a_1$ and $b_1$ to make $c_0$ bounce to $c_1$:
a cooperative sort $ab$ is defined with 4 processes (one for each sub-state of the presence of processes $a_1$ and $b_1$).
For the sake of clarity, the processes of $ab$ are indexed using the sub-state they represent.
Hence, $ab_{01}$ represents the sub-state $\PHstate{a_0,b_1}$, and so on.
Each process of sort $a$ and $b$ hit $ab$ to make it bounce to the process reflecting the status of the sorts $a$ and $b$
(e.g., $\PHfrappe{a_1}{ab_{00}}{ab_{10}}$ and $\PHfrappe{a_1}{ab_{01}}{ab_{11}}$).
Then, to represent the cooperation between $a_1$ and $b_1$, the process $ab_{11}$ hits $c_0$ to make it bounce to $c_1$ instead of independent hits from $b_1$ and $a_1$.

@@

\todo{Nouvelle section pour les hypothèses + conséquences ?}
\subsection{Restrictions} \towrite{Reformuler}
\label{ssec:hypothesis}

In the scope of this paper, we focus on a specific class of PH models.
We define here the restrictions that lead to this class of models \todo{and show that they are equivalent to discrete networks}.

\pref{cr:2prio} allows to models two kinds of actions: controllable actions that model the non-determinacy of biological evolutions,
and uncontrollable actions that are used to model non-biological behaviours in the model, that is, the update of cooperative sorts.
\pref{cr:bounded} states that the dynamics of the studied model $\PH$ contains no infinite sequence of uncontrollable actions.
As these actions can be considered as non-biological and therefore instantaneous, we thus prevent the existence of any Zeno behaviour
which would allow the play of an infinite sequence of high priority actions in zero time.
\begin{criterion}[2 priorities]
\label{cr:2prio}
%\paragraph{2 priorities}
  In this paper, we only consider Process Hitting with $2$ priorities.
\end{criterion}
%
\begin{criterion}[Bounded termination]
\label{cr:bounded}
%\paragraph{Bounded termination}
  The dynamics of $\restriction{\PH}{1}$ contains no cycles:
  %Cycles are still allowed in $\PH$ however.
  %Formally :
  $$\exists N \in \sN, \forall \in \PHl, \forall \delta \in \restriction{\Sce}{1}(s), |\delta| \leq N \enspace.$$
\end{criterion}

In \pref{def:component} we define a well-formed component as a sort that is hit only by actions of priority $2$.
A well-formed component can stand for a real biological component (protein, light, nutriment…).
\begin{definition}
\label{def:component}
  A sort $a \in \PHs$ is a \emph{well-formed component} if and only if:
  $$\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 2 \enspace.$$
  We call $\components$ the set of well-formed components.
\end{definition}

\pref{def:chainsactions} defines chains of uncontrollable actions,
and \pref{cr:cyclefreeness} prevents the presence of cycles in these chains to avoid incoherent states between cooperative sorts.
\begin{definition}
\label{def:chainsactions}
  The set of chains of actions $\PHh(a, b)$ between two sorts $a, b \in \PHs$ is defined as below:
  \begin{align*}
  \PHh(a, b) = \{ (h_i)_{i \in \segm{0}{s}} \in (\PHh^{(1)})^{s+1} &\mid s \in \sN, \PHsort(\PHhitter(h_0)) = a \wedge \PHsort(\PHtarget(h_s)) = b \\
    & \wedge \forall i \in \segm{1}{s}, \PHsort(\PHtarget(h_{i-1})) = \PHsort(\PHhitter(h_{i})) \}
  \end{align*}
\end{definition}
\begin{criterion}[Cycle-freeness between actions of priority $1$]
\label{cr:cyclefreeness}
%\paragraph{Cycle-freeness between actions of priority $1$}
  There is no cycle in chains of actions:
  $$\forall a \in \PHs, \PHh(a, a) = \emptyset \enspace.$$
\end{criterion}

In \pref{def:neighbours} we define the notion of neighbour \towrite{revoir terminologie : neighbors ? vicinity ?} sorts and actions of a sort.
The set of neighbour sorts $\Vs(a)$ of a given sort $a$ is the set of components that can interact with it through actions of priority $1$.
The set of neighbour actions $\Vh(a)$ of $a$ is the set of actions of priority $1$ influencing $a$.
These definitions are permitted by the previous restrictions (\pref{cr:2prio} and~\ref{cr:cyclefreeness}).
\begin{definition}[$\Vs : \PHs \rightarrow \powerset(\PHsort)$ ; $\Vh : \PHs \rightarrow \powerset(\PHh)$]
\label{def:neighbours}
  For all sort $a \in \PHs$,
  \begin{align*}
    \Vs(a) &= \{ b \in \components \mid \PHh(b, a) \neq \emptyset \}\\
    \Vh(a) &= \{ h \in \PHh^{(1)} \mid \exists b \in \Vs(a), \exists hs \in \PHh(b,a), h \in hs \}
  \end{align*}
\end{definition}

Finally, we introduce the notion of local steady-state of a sort from a given state.
This local steady-state is the set of processes towards which the sort tends to evolve to and stay in by playing only uncontrollable actions.
We denote $\pfp_s(a)$ this set of processes for a sort $a$ in state $s$ (\pref{def:pfp}),
and we derive from this the notion of well-formed cooperative sort (\pref{def:cs}).
A well-formed cooperative sort allows to define a cooperation between sorts as presented in \pref{ssec:cooperation};
therefore, it must be only hit by actions of high priority
and in a way that it efficiently represents any configuration of its neighbour sorts.
%any sub-state of its neighbourhood is represented by exactly one local steady state in this sort,
%and all of its processes represents at least one sub-state of its neighbourhood. \towrite{reformuler ou supprimer}
\begin{definition}[$\pfp : \PHl \times \PHs \rightarrow \powerset(\PHproc)$]
\label{def:pfp}
  For all $s \in \PHl$ and $a \in \PHs$,
  \begin{align*}
    \pfp_s(a) = \{ a_i \in \PHl_a &\mid \exists \delta \in \restriction{\Sce}{1}(s), \PHget{(s \PHplay \delta)}{a} = a_i \\
    & \quad \wedge \forall \PHhit{b_i}{c_j}{c_k} \in \Vh(a), \PHget{(s \PHplay \delta)}{b} \neq b_i \vee \PHget{(s \PHplay \delta)}{c} \neq c_j \}
  \end{align*}
  Of course, if $a \in \components$, then $\pfp_s(a) = \{ \PHget{s}{a} \}$.
\end{definition}
%
\begin{definition}
\label{def:cs}
  A sort $a \in \PHs$ is a \emph{well-formed cooperative} sort if and only if:
  \begin{enumerate}
    \item $\forall h \in \PHh, \PHsort(\PHtarget(h)) = a \Rightarrow \prio(h) = 1$
%    \item $\forall s \in \underset{b \in \Vs(a)}{\times} \PHl_b, |\pfp_s(a)| = 1$
%    \item $\forall a_i \in \PHl_a, \exists s \in \underset{b \in \Vs(a)}{\times} \PHl_b, \pfp_s(a) = \{ a_i \}$
    \item \label{csai} $\forall s \in \PHl, \exists a_i \in \PHl_a, \pfp_s(a) = \{ a_i \}$
    \item \label{css} $\forall a_i \in \PHl_a, \exists s \in \PHl, \pfp_s(a) = \{ a_i \}$
  \end{enumerate}
\end{definition}
We call $\cs$ the set of well-formed cooperative sorts.
Because of (\ref{csai}), we denote in the following: $\pfp_s(a) = a_i$.
Furthermore, because of (\ref{css}), we denote $\csState(a_i)$ the set of sub-states represented by the process $a_i$ of any cooperative sort $a$ (\pref{def:csState}).
\begin{definition}[$\csState : \PHproc \rightarrow \powerset(\PHproc)$]
\label{def:csState}
  If $a \in \cs$ and $a_i \in \PHl_a$, we define:
  $$\csState(a_i) = \{ ps \in \underset{b \in \Vs(a)}{\times} \PHl_b \mid \forall s \in \PHl, \pfp_{s \Cap ps}(a) = a_i \}$$
\end{definition}

In the following we will simply write “component” (resp.~“cooperative sort”) instead of “well-formed component” (resp.~“well-formed cooperative sort”).
The last criterion states that the set of sorts of the considered PH must be divided between components and cooperative sorts.

\begin{criterion}[Components \& cooperative sorts]
\label{cr:compcs}
\towrite{Problème : sortes non frappées dans $\components$ ou $\cs$ ?}
%\paragraph{Components \& cooperative sorts}
  %The studied PH is a PH with 2 priorities, and it is only composed of well-formed components and cooperative sorts:
  $$\PHs = \components \cup \cs \wedge \components \cap \cs = \emptyset$$
\end{criterion}

\todo{The criteria given in this subsection allow to define a class of PH models that share the same behaviour than discrete models.
The demonstration of this statement is found in Annex B.}
